Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-01-27 text/src/article.tex
Results section for installation

% }}}
% {{{ Background
\section*{Background}
 Text for this section.
% }}}
\section*{Implementation} %{{{1

\subsection*{Code and Dependencies} %{{{2

Scaffolder is written in the Ruby programming language, version 1.8.7
\cite{ruby-lang}. Unit tests were implemented during development to maintain
functionality of individual elements of source code. Integration tests were
also written to test the scaffold software interface as a whole. These unit and
integration tests were written using the RSpec and Cucumber libraries
respectively\cite{rspec}.\pb

The Scaffolder source code is documented using the Yard library\cite{yard}.
Unix manual pages for the command line interface were generated using the Ronn
library\cite{ronn}. The manipulation of biological sequences in Scaffolder is
performed using the BioRuby library\cite{bioruby}. A full list of the software
dependencies in scaffolder can be found in the Gemfile in the root of source
code directory.\pb

\subsection*{Scaffold file syntax} %{{{2

The choice of nucleotide sequences comprising the genome scaffold is specified
in YAML \cite{yaml}. YAML is a language format for data using white-space and
indentation to produce a machine readable structure. As a standardised data
format third-party developers have the option to generate a genome scaffold
using any programming language for which a YAML parser exists. Furthermore YAML
formatted scaffold files can be validated using third party tools such as
Kwalify\cite{kwalify}.\pb

The initial assembly of sequencing reads may result in an incomplete genome
build. Comparison of contigs to reference genomes and the generation of
additional sequence from PCR translates to genome finishing as an ongoing
process. Therefore the scaffold file should be simple to edit manually in
addition to being computationally tractable. This requirement for the scaffold
files to be modifiable in a simple simple text editor is the second reason YAML
was selected as the format for scaffold files.\pb

The form of the scaffold file is a list of entries. Each entry corresponds to
a length of sequence in the the draft genome sequence. When generating the
genome sequence from the scaffold file each entry is processed in order. Each
of the nucleotide sequence specified by the scaffold file entries are
concatenated together to produce the draft genome sequence. Each entry in the
scaffold file may have attributes which define whether a sub-sequence or the
reverse complement of the sequence should be used is the scaffold. The types of
attributes are outlined in the Results and Discussion section below.\pb

\section*{Results and Discussion} %{{{1 \subsection*{Results sub-heading}
\subsubsection*{This is a sub-sub-heading} Sub-sub-sub-headings are made with
the \textsl{\\subsubsection} command. \pb pb at end of lines ensures correct
paragraph spacing.\pb Text for this sub-sub-section \ldots
\subsubsection*{Another sub-sub-sub-heading} Text for this sub-sub-section
\ldots

  \subsection*{Another results sub-heading}
    Text for this sub-section \ldots

  \subsection*{Yet another results sub-heading}
    Text for this sub-section.  More results \ldots
% }}}
% {{{ Conclusions
%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots
% }}}
% {{{ Availability and Requirements
\section*{Availability and Requirements}

@@

% }}}
% {{{ Competing Interests
\section*{Authors contributions}
  The authors declare no competing interests.
% }}}
% {{{ Availability and Requirements
\section*{List of Abbreviations Used}

  \begin{description}
    \item[YAML:] YAML Ain't Markup Language\cite{yaml}
  \end{description}

% }}}
% {{{ Author Contributions
\section*{Authors contributions}

>>>>

% }}}
% {{{ Background
\section*{Background}
 Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
% }}}
% {{{ Implementation
%%%%%%%%%%%%%%%%%%%%%%
\section*{Implementation}
  Text for this section \ldots
% }}}
\section*{Results and Discussion} %{{{1

\subsection*{Scaffolder Description} %{{{2

Individual reads produced from nucleotide sequencing can be assembled into
contiguous regions (``contigs'') of many overlapping individual reads. These
reads may however not represent a complete genome sequence where all contigs do
not result in a single complete sequence. The software outlined here
``Scaffolder'' allows these contigs to be further joined together into
a continuous single sequence. This selection, order and direction of contigs
into a final draft genome is often described as the `scaffold'. Scaffolder
allows the manual or computational specification of a genome scaffold by
specifying which contig sequences are used in simple to edit configuration
files. \pb

The scaffold file specifies how contigs should be joined together in the draft
genome sequence. The raw nucleotide sequence for each contig is stored in
a fasta file, as might be produced from assembly of sequencing reads using
programs such as Newbler\cite{miller2010} or ABySS\cite{simpson2009}.
Scaffolder then the contig to be selected by specifying the name of the contig
in the scaffold file. The scaffold file also allows a sub-sequence or the
reverse complement of a contig to used. This therefore preserves the original
sequence data but also allowing sequences to manipulated in the final genome
build. \pb

The majority of a genome can be estimated from scaffolding assembled contigs
into the most parsimonious build. There may however still be internal gaps in
the contigs from mate pair assembly based on estimated mate pair insert size.
Scaffolder allows inserts to be added to scaffolded so that gaps may be bridged
and filled - a common process in finishing a genome sequence. Inserts may be
treated the same way as a contig where the raw sequence, as may be produced by
PCR is stored in a fasta file, and then trimmed and/or reversed to fill the gap
in the contig. \pb

Scaffolder allows the specification of unknown regions in the scaffold. These
are areas of the genome where the nucleotide sequence is unknown but the
approximate size region may be estimated from mate pair insert size or from
mapping contigs to reference genome and estimating the size of the gap. These
unknown regions allow a continuous draft sequence to be scaffolded but whilst
still highlighting regions waiting to be resolved. \pb

\subsection*{Defining a scaffold in a text file} %{{{2

The genome scaffold is wriiten in YAML format. YAML is a data format designed
to be both machine readable and easy to manually edit. An example of this
format illustrating all the features of scaffolder is shown in Figure 1. \pb

\subsubsection*{Simple sequence region} %{{{3

The first line of the file begins with three dashes. This is required to
signify the start of a YAML document. Lines 2 begins with a dash character `-'
which denotes the start of an entry in the sequence scaffold, in this case the
first entry. Line 3 is indented by two spaces which therefore means this line
contains an entry type. The ``sequence'' tag is used indicate a sequence is
inserted at this region in the scaffold. The following line contain is
indented by two spaces and indicates that this an attribute of the sequence
entry. In this case the ``source'' keyword is used to indicate used which
sequence should be inserted into the scaffold. This sequence is indentified by
the first word of the fasta header of the selected sequence. This indicates
a simple sequence region added to the scaffold. \pb

\subsubsection*{Unresolved seqeunce region} %{{{3

The following line has no indentation and begins with dash indicating the next
entry in the scaffold. The following entry is indentified by the ``unresolved''
tag. This identifies a region of unknown nucleotide sequence but is included
because the approximate size of the region is known. The following line defines
that the size of this unknown region in the genome scaffold should be 20
nucleotides. This therefore results in the region being filled with 20 `N'
characters in draft sequence. \pb

\subsubsection*{Trimmed sequence region with multiple inserts} %{{{3

The following entry is another sequence using the sequence from 'sequence2' in
the scaffold file. This sequence entry has three further attributes in addition
to defining the source sequence. The first two are start and stop. These are
used to trim the final sequence inserted into the scaffold. In this case the
first two nucleotides of the sequence are removed so that that sequence starts
at position 3 of the original raw sequence. Similarly the sequence after
position is trimmed away after position 100. These sequence is also reverse
complemented using the ``reverse'' tag. Reverse complementation is performed
after the sequence has been trimmed to the specified coordinates. \pb

The final attribute for this sequence entry defines inserts to be added to the
sequence. These are additional regions of sequence that can be added to
a sequence entry to alter the existing sequence such to fill gaps or correct
sequencing errors. Inserts are defined using the ``inserts'' tag. Each insert
is then listed similarly to the entries in the scaffold: beginning with a dash
on one line then indented by an additional two spaces on the following lines.
\pb

Inserts are listed similarly to sequence entries where reverse, start and stop
attributes can be used to trim and reverse the sequence. The `source' tag is
used to indentify which fasta sequence entry should be used as the insert.  The
insert has two additional tags ``open'' and ``close'' which determine where the
insert is added to the hosting sequence. These coordinates are inclusive
replaceing the coordinate nucleotides. The open and close positions need not be
the same size as the insert size. Any difference in size to the replaced
nucleotides will result in the length on the encoding sequence being adjusted.
\pb

The next insert entry shows an insert with only the open position specified.
This is to illustrate that only the one of the `open' or `close' keywords is
sufficient to add an insert. If only one of these words is used then the
opposing insert position is calculated from the length on the insert sequence
in the fasta file. This feature is designed for using inserts to close gaps
where the length of the insert may not completely fill the gap but can still be
used to fill a portion of the gap sequence. \pb

\subsubsection*{Scaffolder file attribute processing} %{{{3

Each scaffolder sequence entry can have multiple attributes which applied in
different order can result in a different sequence. For instance reverse
complementing the sequence then adding inserts will have a different effect
compared with performing these actions in the reverse order. Scaffolder
therefore follows the following protocol for processing each entry in the
scaffold file. \pb

\paragraph{Reverse sort inserts:} If the entry is a sequence and has sequence
inserts these are reverse sorted by insert close position. The insert with the
largest close position is first and and the smallest close position is last.
This is done to preserve the the insert open/close coordinates as they are
added to the scaffold. \pb

\paragraph{Update sequence with insert:} Each insert is added to the sequence
replacing the region specified by the open an close position with the specified
source sequence.\pb

\paragraph{Update sequence start/stop co-ordinates:} If the insert added to the
sequence is a different length than the region replaced the start and stop
co-ordinates of the sequence are correspondingly updated if necessary. E.g. if
a 7 bp insert filled a bp region this would result in a 2bp difference in the
length of the sequence. The stop coordinated would be updated to reflect this.
\pb

\paragraph{Sub-sequence selected:} If the sequence is has start or stop
sequences specified these are used to trip the sequence to these co-ordinates.
\pb

\paragraph{Sub-sequence reverse complemented: } If specified with the reverse keyworkd the sequence is reverse complemented.
\pb

\subsection*{Scaffolder usage} %{{{2

Scaffolder provides a ruby API that allows the scaffold and designated
sequences to be manipulated through a ruby programme. This allows scaffolder to
be integrated into exiting genomics works flows such as using Rake.  Scaffolder
all provides a command line binary for the manipulation of the scaffold. This
binary behaves as a standard Unix command line interface returning appropriate
exit statuses and providing manual pages for each of the commands. These two
methods are outlined in the following three sections. \pb

\subsubsection*{Installation} %{{{3

Scaffolder relies on both the ruby programming language and the RubyGems
package management system. These are commonly installed on recent versions of
Mac OSX and linux distrubtions. The installation of these can be tested on the
command line by typing the following commands. \pb

\begin{verbatim}
ruby -v
gem -v
\end{verbatim}

If both these commands return without error this indicates the necessary
requirements to install scaffolder. The scaffolder API and command line tools
can then be installed with the follow command.

\begin{verbatim}
gem install scaffolder scaffolder-tools
\end{verbatim}

Finally to ensure that the command line tools are accessible these the
RubyGems binary directory should be in the command line path. This can be done
by adding the following to the ~/.bashrc file or other corresponding shell
configure file. Assuming that ~/.gem is the install path for RubyGems.

\begin{verbatim}
export PATH=~/.gem/bin:$PATH
\end{verbatim}

\subsubsection*{Command line interface} %{{{3

  * Illustrate running command line to generate output
  * Demonstrate scaffold validate to test inserts overlap

\subsubsection*{Accessible programmatic interface} %{{{3

  * Demonstrate how scaffolder can be manipulated programmatically
  * Examine scaffolder data using a ruby program
  * Example of extending sequence class with a new method \%GC content
  * Create a new class to do something

\subsection*{Discussion of scaffolder approach} %{{{2

\section*{Conclusions} %{{{1

Text for this section \ldots

\section*{Availability and Requirements}

@@

% }}}
% {{{ Competing Interests
\section*{Authors contributions}
  The authors declare no competing interests.
% }}}
% {{{ Author Contributions
\section*{Authors contributions}
